### 2020-02-28 thesis/content/chapter-related-work.tex
continued sgt section

In order to process a graph $G$ of size $n = |\mathcal{V}_G|$, one is forced to choose some encoding, e.g.\ an adjacency matrix $A \in {\{0, 1\}}^{n \times n}$.
Such an encoding introduces a vertex ordering $v_1, \dots, v_n$ that does not carry any semantic meaning.
Consequently there are $n!$ equivalent encodings of $G$. % chktex 40
To represent those encodings we introduce the notion of ordered induced subgraphs.
\begin{defn}\label{defn:related:ordered-subgraph}
	For all vertex $k$-tuples $v = (v_1, \dots, v_k) \in \mathcal{V}_G^k$ let $\hat{v} = \{ v_i\, |\, i \in [k] \}$.
	Then $G[v] \coloneqq (\hat{v}, \mathcal{E}_G \cap \hat{v}^2, v)$ is called the ordered subgraph of $G$ induced by $v$.
	If $\hat{v} = \mathcal{V}_G$ and $k = |\mathcal{V}_G|$, we call $v$ an ordering of $G$ and $G[v]$ an encoding of $G$.
\end{defn}
\begin{defn}\label{defn:related:ordered-subgraph-equivalence}
	Two ordered induced subgraphs $G[v]$ and $H[w]$ with $(v_1, \dots, v_k) \in \mathcal{V}_G^k$ and $(w_1, \dots, w_k) \in \mathcal{V}_H^k$ are \textit{equivalent} ($G[v] \equiv H[w]$) iff.\
	\begin{align*}
		&\quad\forall i, j \in [k]: (v_i, v_j) \in \mathcal{E}_G \leftrightarrow (w_i, w_j) \in \mathcal{E}_H \\
		&\, \begin{aligned}
			\land&\ \forall i, j \in [k]: l_G[v_i, v_j] = l_H[w_i, w_j] &\land&\ \forall i \in [k]: l_G[v_i] = l_H[w_i] \\
			\land&\ \forall i, j \in [k]: x_G[v_i, v_j] = x_H[w_i, w_j] &\land&\ \forall i \in [k]: x_G[v_i] = x_H[w_i]\text{.}
		\end{aligned}
	\end{align*}
	Using this notion of equivalence, we call $[G[v]] \coloneqq \{ G[w]\, |\, G[w] \equiv G[v] \land w \in \mathcal{V}_G^{*} \}$ the ordered subgraph equivalence class of $G[v]$.
\end{defn}
\begin{defn}
	Two graphs $G$ and $H$ are \textit{isomorphic} ($G \simeq H$) iff.\ there are equivalent encodings $G[v] \equiv H[w]$ of them.
	Consequently $[G] \coloneqq \{ H\, |\, H \simeq G \}$ is called the isomorphism class of $G$.
\end{defn}
The goal of the \ac{gi} problem is to check whether $G \simeq H$ for two arbitrary graphs.
Even though there is no known universal algorithm in $P$ that solves \ac{gi}, \citet{Babai1980} showed that almost all graphs can be trivially distinguished in linear time.
More recently \citet{Babai2015} also presented a quasipolynomial upper time bound for the remaining hard \ac{gi} instances.
For all practical purposes, \ac{gi} can therefore be solved efficiently; e.g.\ via the \texttt{nauty} program~\cite{McKay}\cite{McKay2013}.
One important subroutine in most \ac{gi} checkers is the \acl{wl} algorithm which will be described next.

@@

If we increase the \acs{wl}-dimension to $k = 2$, the described 1-\acs{wl} restrictions no longer apply.
2-\acs{wl} is able to distinguish more than $1 - \mathcal{O}(\nicefrac{1}{n})$ of the regular $n$-vertex graphs~\cite[cor.~1.8.6]{Immerman1990} and can even count cycles.
\begin{prop}[see \citet{Fuerer2017} and \citet{Arvind2019}]\label{prop:related:wl2-cycle-count}
	2-\acs{wl} is able to count $m$-cycles for $m \leq 7$ but it cannot detect 8-cycles.
\end{prop}
\begin{hproof}
	Only the idea behind triangle and 4-cycle counting is outlined here to give an intuition for why \cref{prop:related:wl2-cycle-count} holds.
	Note that the 2-\acs{wl} neighbors of each edge $(v, w) \in \mathcal{E}_G$ correspond to all possible paths $(v, u, w)$ of length 2, i.e.\ all possible triangles.
	By \cref{defn:related:wlk-refine} there must thus be a color subset $C^{\triangle}_j \subseteq \mathcal{C}$ representing that an edge is involved in exactly $j$ triangles after one refinement step.
	2-\acs{wl} then trivially counts triangles by setting $g(\mathit{dist}_\chi) = \frac{1}{3} \sum_j j \sum_{c \in C^{\triangle}_j} \mathit{dist}_\chi(c)$ to satisfy \cref{defn:related:wl-compute}.
	Analogously for 4-cycles, let $C^{\square}_j \subseteq \mathcal{C}$ be the colors indicating a non-edge $(v, w) \notin \mathcal{E}_G$ with $j$ common vertex neighbors $u \in \Gamma_G(v) \cap \Gamma_G(w)$.
	The number of 4-cycles is then determined by the colors of the diagonals through them via $g(\mathit{dist}_\chi) = \frac{1}{2} \sum_{j \geq 2} \binom{j}{2} \sum_{c \in C^{\square}_j} \mathit{dist}_\chi(c)$.
	Using a similar but more involved combinatorial argument requiring multiple color refinement steps, 5-, 6- and 7-cycle counting can be shown.
\end{hproof}
As we just saw, 2-\acs{wl} is significantly more powerful than 1-\acs{wl}.
By \cref{prop:related:wl-upper} there are of course still 2-\acs{wl} indistinguishable graphs;
among others those are the strongly regular graphs $\mathit{srg}_{n, d, \lambda, \mu}(G)$.
\begin{defn}
	$\begin{aligned}[t]
		\mathit{srg}_{n, d, \lambda, \mu}(G) \iff\ & {|\mathcal{V}_G|} = n \quad\land\quad \forall v \in \mathcal{V}_G: {|\Gamma_G(v)|} = d \\
		&\land \forall (v, w) \in \mathcal{E}_G: {|\Gamma_G(v) \cap \Gamma_G(w)|} = \lambda \\
		&\land \forall (v, w) \in \mathcal{V}_G^2 \setminus \mathcal{E}_G: {|\Gamma_G(v) \cap \Gamma_G(w)|} = \mu
	\end{aligned}$
\end{defn}
Generally this restriction of 2-\acs{wl} is not an issue since strongly regular graphs do not appear in typical \ac{gcr} datasets.
Lastly there are few results regarding the classes of distinguishable graphs and computable functions for higher \ac{wl}-dimensions.
For $k = 3$, \citet{Kiefer2017} have shown that all planar graphs can be distinguished.

@@

An alternative perspective on graph characterization is provided by \textit{spectral graph theory}.

>>>>

In order to process a graph $G$ of size $n = |\mathcal{V}_G|$, one is forced to choose some encoding, e.g.\ an adjacency matrix $A_G \in {\{0, 1\}}^{n \times n}$.
Such an encoding introduces a vertex ordering $v_1, \dots, v_n$ that does not carry any semantic meaning.
Consequently there are $n!$ equivalent encodings of $G$. % chktex 40
To represent those encodings we introduce the notion of ordered induced subgraphs.
\begin{defn}\label{defn:related:ordered-subgraph}
	For all vertex $k$-tuples $v = (v_1, \dots, v_k) \in \mathcal{V}_G^k$ let $\hat{v} = \{ v_i\, |\, i \in [k] \}$.
	Then $G[v] \coloneqq (\hat{v}, \mathcal{E}_G \cap \hat{v}^2, v)$ is called the ordered subgraph of $G$ induced by $v$.
	If $\hat{v} = \mathcal{V}_G$ and $k = |\mathcal{V}_G|$, we call $v$ an ordering of $G$ and $G[v]$ an encoding of $G$.
\end{defn}
\begin{defn}\label{defn:related:ordered-subgraph-equivalence}
	Two ordered induced subgraphs $G[v]$ and $H[w]$ with $(v_1, \dots, v_k) \in \mathcal{V}_G^k$ and $(w_1, \dots, w_k) \in \mathcal{V}_H^k$ are \textit{equivalent} ($G[v] \equiv H[w]$) iff.\
	\begin{align*}
		&\quad\forall i, j \in [k]: (v_i, v_j) \in \mathcal{E}_G \leftrightarrow (w_i, w_j) \in \mathcal{E}_H \\
		&\, \begin{aligned}
			\land&\ \forall i, j \in [k]: l_G[v_i, v_j] = l_H[w_i, w_j] &\land&\ \forall i \in [k]: l_G[v_i] = l_H[w_i] \\
			\land&\ \forall i, j \in [k]: x_G[v_i, v_j] = x_H[w_i, w_j] &\land&\ \forall i \in [k]: x_G[v_i] = x_H[w_i]\text{.}
		\end{aligned}
	\end{align*}
	Using this notion of equivalence, we call $[G[v]] \coloneqq \{ G[w]\, |\, G[w] \equiv G[v] \land w \in \mathcal{V}_G^{*} \}$ the ordered subgraph equivalence class of $G[v]$.
\end{defn}
\begin{defn}
	Two graphs $G$ and $H$ are \textit{isomorphic} ($G \simeq H$) iff.\ there are equivalent encodings $G[v] \equiv H[w]$ of them.
	Consequently $[G] \coloneqq \{ H\, |\, H \simeq G \}$ is called the isomorphism class of $G$.
\end{defn}
The goal of the \ac{gi} problem is to check whether $G \simeq H$ for two arbitrary graphs.
Even though there is no known universal algorithm in $P$ that solves \ac{gi}, \citet{Babai1980} showed that almost all graphs can be trivially distinguished in linear time.
More recently \citet{Babai2015} also presented a quasipolynomial upper time bound for the remaining hard \ac{gi} instances.
For all practical purposes, \ac{gi} can therefore be solved efficiently; e.g.\ via the \texttt{nauty} program~\cite{McKay}\cite{McKay2013}.
One important subroutine in most \ac{gi} checkers is the \acl{wl} algorithm which will be described next.

@@

If we increase the \acs{wl}-dimension to $k = 2$, the described 1-\acs{wl} restrictions no longer apply.
2-\acs{wl} is able to distinguish more than $1 - \mathcal{O}(\nicefrac{1}{n})$ of the regular $n$-vertex graphs~\cite[cor.~1.8.6]{Immerman1990} and can even count cycles:
\begin{prop}[see \citet{Fuerer2017} and \citet{Arvind2019}]\label{prop:related:wl2-cycle-count}
	2-\acs{wl} is able to count $m$-cycles for $m \leq 7$ but it cannot detect 8-cycles.
\end{prop}
\begin{hproof}
	Only the idea behind triangle and 4-cycle counting is outlined here to give an intuition for why \cref{prop:related:wl2-cycle-count} holds.
	Note that the 2-\acs{wl} neighbors of each edge $(v, w) \in \mathcal{E}_G$ correspond to all possible paths $(v, u, w)$ of length 2, i.e.\ all possible triangles.
	By \cref{defn:related:wlk-refine} there must thus be a color subset $C^{\triangle}_j \subseteq \mathcal{C}$ representing that an edge is involved in exactly $j$ triangles after one refinement step.
	2-\acs{wl} then trivially counts triangles by setting $g(\mathit{dist}_\chi) = \frac{1}{3} \sum_j j \sum_{c \in C^{\triangle}_j} \mathit{dist}_\chi(c)$ to satisfy \cref{defn:related:wl-compute}.
	Analogously for 4-cycles, let $C^{\square}_j \subseteq \mathcal{C}$ be the colors indicating a non-edge $(v, w) \notin \mathcal{E}_G$ with $j$ common vertex neighbors $u \in \Gamma_G(v) \cap \Gamma_G(w)$.
	The number of 4-cycles is then determined by the colors of the diagonals through them via $g(\mathit{dist}_\chi) = \frac{1}{2} \sum_{j \geq 2} \binom{j}{2} \sum_{c \in C^{\square}_j} \mathit{dist}_\chi(c)$.
	Using a similar but more involved combinatorial argument requiring multiple color refinement steps, 5-, 6- \& 7-cycle counting can be shown.
\end{hproof}
As we just saw, 2-\acs{wl} is significantly more powerful than 1-\acs{wl}.
Though, by \cref{prop:related:wl-upper}, there are of course still 2-\acs{wl} indistinguishable graphs;
among others those are the strongly regular graphs $\mathit{srg}_{n, d, \lambda, \mu}(G)$.
\begin{defn}
	$\begin{aligned}[t]
		\mathit{srg}_{n, d, \lambda, \mu}(G) \iff\ & {|\mathcal{V}_G|} = n \quad\land\quad \forall v \in \mathcal{V}_G: {|\Gamma_G(v)|} = d \\
		&\land \forall (v, w) \in \mathcal{E}_G: {|\Gamma_G(v) \cap \Gamma_G(w)|} = \lambda \\
		&\land \forall (v, w) \in \mathcal{V}_G^2 \setminus \mathcal{E}_G: {|\Gamma_G(v) \cap \Gamma_G(w)|} = \mu
	\end{aligned}$
\end{defn}
Generally this restriction of 2-\acs{wl} is not an issue since strongly regular graphs do not appear in typical \ac{gcr} datasets.
By going to $k = 3$, even some strongly regular graphs as well as all planar graphs can be distinguished though~\cite{Kiefer2017}.
Apart from that there are currently few results regarding the classes of distinguishable graphs and computable functions for even higher \ac{wl}-dimensions.

@@

Let us now look at an alternative perspective on graph characterization which is provided by \textit{spectral graph theory}.
While \ac{wl} characterizes a graph via its color distribution $\mathit{dist}_{\chi_{G,k}^{*}}$, the spectral approach uses its so-called \textit{spectrum} $\lambda_G = (\lambda_{G, 1}, \dots, \lambda_{G, n}) \in \mathbb{R}^n$ where $n = |\mathcal{V}_G|$.
The key idea behind this is to interpret the adjacency matrix $A_G \in \mathbb{R}^{n \times n}$ of $G$ not simply as an encoding of the edges ($A_{G,i,j} = \mathbbm{1}[(v_i, v_j) \in \mathcal{E}_G]$) but as a linear operator $A_G: (\mathcal{V}_G \to \mathbb{R}) \to (\mathcal{V}_G \to \mathbb{R})$ acting on the vector space of real-valued functions with a vertex domain.
This means that $A_G$ transforms so-called \textit{graph signals} $x: \mathcal{V}_G \to \mathbb{R}$ which are functions assigning \textit{signal strengths} to vertices.
By applying $A_G x$ the signal strength $x[v_i]$ of each vertex is added to the signal strengths of its neighbors $\Gamma_G(v_i)$.
Using this functional perspective, one can characterize graphs via the \acf{ft}.
We will now see how this is done;
however, since a comprehensive description of spectral graph theory would exceed the scope of this thesis, only a brief overview will be given.
For a more detailed introduction to the field we refer to \citet{Shuman2013}.

\paragraph{\ac{ft} on real domains}
Let us begin by describing the \ac{ft} for the more common case of functions with real domains.
All functions $f: \mathbb{R} \to \mathbb{R}$ can be interpreted as infinite-dimensional vectors $f = \int_{\mathbb{R}} f(t)\, b_t \dif t$ with $b_t: \mathbb{R} \to \mathbb{R}$ being a standard basis vector/function defined as $b_t(s) \coloneqq \begin{cases} 1 & \text{if } s = t \\[-6pt] 0 & \text{else} \end{cases}$. % chktex 21
The values of $f$ are then described as its $b_t$ components $\langle b_t, f \rangle = f(t)$, where $\langle b_t, \cdot \rangle = \delta_t(\cdot)$ denotes the Dirac delta function translated by $t$.
The \acl{ft} $\hat{f}$ of $f$ corresponds to a change of basis from the standard basis vectors $b_t$ to the Fourier basis vectors $u_\xi(t) \coloneqq e^{2 \pi i \xi t}$, i.e.\ $f = \int_{\mathbb{R}} \hat{f}(t)\, u_\xi \dif \xi$.
The Fourier basis is characterized by the fact that it is an eigenbasis of the so-called \textit{Laplace operator} $\Delta$, i.e.\ $\Delta u_\xi = \lambda_\xi u_\xi$ with $\lambda_\xi$ being the eigenvalue corresponding to $u_\xi$.
In the real domain this Laplace operator $\Delta$ is effectively the same as the second-derivative $\frac{\dif^2}{\dif t^2}$.
Thus it is easy to see that indeed $\Delta u_\xi = \frac{\dif^2}{\dif t^2} u_\xi = \lambda_\xi u_\xi$ for $\lambda_\xi = -{(2 \pi \xi)}^2$.

\paragraph{\ac{ft} on vertex domains}
To extend the notion of the \ac{ft} from real functions $f: \mathbb{R} \to \mathbb{R}$ to graph signals $x: \mathcal{V}_G \to \mathbb{R}$ we need to define a graph variant of the Laplace operator $\Delta$.
It turns out that there are multiple possible ways to do so, the most simple being the so-called \textit{combinatorial graph Laplacian}.
\begin{defn}
	The combinatorial graph Laplacian $L_G \in \mathbb{R}^{n \times n}$ is defined as
	\begin{align*}
		L_G \coloneqq D_G - A_G\quad\text{with the degree matrix } D_G\coloneqq {
			\renewcommand*{\arraystretch}{0.5}
			\begin{pmatrix}
				d_1 & & \\
				& \ddots & \\
				& & d_n
			\end{pmatrix}
		}, d_i \coloneqq |\Gamma_G(v_i)| \text{.}
	\end{align*}
\end{defn}
Using this definition, applying $L_G x$ is analogous to taking the second derivative $\frac{\dif^2}{\dif t^2} f$.
Putting both Laplacian variants, $L_G$ and $\frac{\dif^2}{\dif t^2}$, side-by-side gives an intuition for why this is the case:
\begin{equation*}
	\begin{split}
		-\frac{\dif^2}{\dif t^2} f(t) = \smashoperator[l]{\lim_{h \to 0}} \frac{1}{h^2} (\underbrace{f(t) - f(t-h)}_{\Delta_{t, t-h}} + \underbrace{f(t) -  f(t+h)}_{\Delta_{t, t+h}})
	\end{split}
	\begin{split}\ \left|\ %
		L_G x[v] = \smashoperator[l]{\sum_{u \in \Gamma_G(v)}}\mkern-6mu \underbrace{(x[v] - x[u])}_{\Delta_{v, u}}
	\right.\end{split}
\end{equation*}
The second derivative of a function $f$ essentially averages the value differences in the neighborhood of a point $t$.
For real-valued functions this neighborhood only consists of the two infinitesimally close points to the left and to the right of $t$, i.e.\ $t - h$ and $t + h$.
The combinatorial graph Laplacian represents the same operation, where each point/vertex $v$ might however have more than two neighbors $u \in \Gamma_G(v)$ that need to be averaged.

The \ac{ft} of a graph signal $x$ then is $\hat{x}[i] = \langle u_{G,i}, x \rangle$ with $u_G = {\{ u_{G,i} \}}_{i=1}^{n}$ being the eigenvectors of $L_G$, s.t.\ $x = \sum_{i=1}^n \hat{x}[i] u_{G,i}$.
Note that $L_G$ is guaranteed to only have real eigenvectors $u_G$ and eigenvalues $\lambda_G$ due to the fact that we only consider undirected graphs with symmetric adjacency matrices $A_G$.
By convention we assume that the eigenvalues are ordered ascendingly: $\lambda_{G,0} \leq \cdots \leq \lambda_{G,n}$.
Those eigenvalues are called the \textit{spectrum of $G$} and they are a vertex-permutation-invariant graph characterization.
Intuitively those eigenvalues describe the connectivity between different parts of the graph.
The initial eigenvalues represent connectivity at a general, coarse level while the last eigenvalues represent the connectivity between finer substructures.
\Cref{fig:related:graph-fourier} illustrates this idea.
A more concrete example for the relation between a graph's structure and its spectrum is that a graph with $m$ connected components has exactly $m$ zero eigenvalues, i.e.\ $\lambda_{G,1} \cdots = \lambda_{G,m} = 0$.
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/related-work/graph-fourier.pdf}
	\caption[Comparison between the basis vectors of the real domain \acl{ft} and the graph \acl{ft}.]{
		Comparison between the basis functions/vectors of the real domain \ac{ft} and the graph \ac{ft}.
		For the eigenfunctions on the upper half only the real cosine components of the complex exponentials are shown.
	}\label{fig:related:graph-fourier}
\end{figure}
